Non-convergence in iterative imputation

### 2025-05-02 Manuscript/manuscript.qmd
update identifiers in background section

Multiple imputation has become the de facto standard to accommodate missing data in social scientific research. The technique enables researchers to draw valid inferences from incomplete data [@rubin76]. [TODO: remove the next sentence?] This is in contrast to ad hoc solutions such as list-wise deletion, which may bias scientific estimates. Multiple imputation separates the missing data problem from the scientific problem. First, the missing values are imputed (i.e., filled in) using an imputation algorithm. Once the data are completed, the analysis of scientific interest can be performed on the completed data. This process is repeated several times to obtain a distribution of plausible values, which reflects the uncertainty in the data due to missingness. Finally, the scientific estimates are pooled across the multiple imputations. This pooled estimate is unbiased and confidence-valid if--and only if--both the missing data problem and the scientific problem are appropriately considered.

[TODO: add refs? e.g., rubin after 18 years paper? murray 2018?]

To obtain valid scientific estimates from incomplete data, both the missing data problem and the scientific problem should be appropriately considered. The popular imputation algorithm 'MICE' has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buuren2011; @buur18]. The algorithm is named after its iterative nature: missing data are imputed on a variable-by-variable basis, using a sequence of univariate imputation models. The validity of this iterative process naturally depends on the convergence of the imputation algorithm. Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.

@@

Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation. Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm produces some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. A widely accepted practice is visual inspection of the algorithm [@ragh07; @buur18]. Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. [add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p. 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred. Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been evaluated on imputation algorithms.

[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]
<!-- It is challenging to arrive upon a single point at which convergence has been reached. -->

@@

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@buuren2011]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values on a variable-by-variable basis. The algorithm is initialized by filling in $m$ sets of starting values, randomly drawn from the observed data. The algorithm then iterates over the $p$ variables, generating $m$ sets of imputed values for each variable $j$. After one full run through the data, the imputed values will be 'updated' in the second iteration. This process should be repeated until convergence is reached.

@@

Iterative imputation algorithms such as MICE often are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].

@@

In 2019, Vehtari et al. proposed an updated version of the potential scale reduction factor $\widehat{R}$, originally coined in 1992. The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity. This version uses three transformations on the scalar summary $\theta$, before computing $\widehat{R}$-values. Namely, rank-normalization, folding, and localization. We follow Vehtari et al.'s formulation (2019, p. 5) to define $\widehat{R}$. Let $m$ be the total number of chains, $\tau$ the number of iterations per chain (where $\tau\geq2$), and $\theta$ the scalar summary of interest. For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$. [TODO: check this! isn't it because of under-dispersion and reliance on observed data?] Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim, therefore, is to reach approximate convergence [@gelm13]. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.
<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->

@@

[TODO: add guaranteed convergence because compatible imputation models, equivalent to JM and true Gibbs sampler, @vanbuuren2007]

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\mathrm{sim}} = 2000$). To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$). We differentiate between six different missingness scenarios that are defined in the data generating mechanism (see Table XYZ).

@@

We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice]. On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest. Our estimands are the regression coefficients $\boldsymbol{\beta}$,

>>>>

Multiple imputation has become the de facto standard to accommodate missing data in social scientific research. The technique enables researchers to draw valid inferences from incomplete data [@rubin76]. Multiple imputation separates the missing data problem from the scientific problem. First, the missing values are imputed (i.e., filled in) using an imputation algorithm. Once the data are completed, the analysis of scientific interest can be performed on the completed data. This process is repeated several times to obtain a distribution of plausible values, which reflects the uncertainty in the data due to missingness. The analysis of scientific interest thus yields one estimate for each imputed dataset. Afterwards, the estimates from the different imputations are pooled into a single estimate. This pooled estimate is unbiased and confidence-valid if--and only if--both the missing data problem and the scientific problem are appropriately considered.

<!-- [TODO: add refs? e.g., rubin after 18 years paper? murray 2018?] -->
<!--  This is in contrast to ad hoc solutions such as list-wise deletion, which may bias scientific estimates. -->

To obtain valid scientific estimates from incomplete data, both the missing data problem and the scientific problem should be appropriately considered. The popular imputation algorithm 'Multivariate Imputation by Chained Equetions' (MICE) has proven to be a powerful tool to draw valid inferences from incomplete data under many circumstances [@buuren2011; @buur18]. The algorithm is named after its iterative nature: missing data are imputed on a variable-by-variable basis, using a sequence of univariate imputation models. The validity of this iterative process naturally depends on the convergence of the imputation algorithm. Yet, there has not been a systematic study on how to evaluate the convergence of iterative imputation algorithms.

@@

Determining whether an algorithm has converged is not trivial, especially in the context of iterative imputation. Since the aim of iterative imputation is to converge to a distribution and not to a single point, the algorithm produces some fluctuations even after it has converged. Because of this property, it may be more desirable to focus on *non*-convergence. A widely accepted practice is visual inspection of the algorithm [@ragh07; @buur18]. Diagnosing non-convergence through visual inspection may, however, be undesirable: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18]. [TODO: add argument: "Inspection of such plots is a notoriously unreliable method of assessing convergence and in addition is unwieldy when monitoring a large number of quantities of interest, such as can arise in complicated hierarchical models" (Gelman et al., 2013, p. 285).] Therefore, a quantitative diagnostic method to assess convergence would be preferred. Fortunately, there are non-convergence identifiers for *other* iterative algorithms, but the validity of these identifiers has not been evaluated on imputation algorithms.

<!-- It is challenging to arrive upon a single point at which convergence has been reached. -->

@@

[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets. On each of these datasets, the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE' [@buuren2011]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values on a variable-by-variable basis. The algorithm is initialized by filling in $m$ sets of starting values, randomly drawn from the observed data. The algorithm then iterates over the $p$ variables, generating $m$ sets of imputed values for each variable $j$. After one full run through the data, the imputed values are 'updated' in the next iteration. This process should be repeated until convergence is reached.

@@

Iterative imputation algorithms such as MICE often are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm will vary even after convergence [@gelm13]. The most converged state that such an algorithm may reach is *approximate*  convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09].

@@

In 2021, Vehtari et al. published an updated version of the potential scale reduction factor $\widehat{R}$, originally coined by Gelman and Rubin in 1992. This adapted version aims to detect non-mixing even in the tails of distributions, and non-stationarity over iterationsuses. To achieve this, the 2021 version uses three transformations on the scalar summary $\theta$ before computing $\widehat{R}$-values: rank-normalization, folding, and localization. To define $\widehat{R}$, we largely follow Vehtari et al.'s formulation (2019, p. 5), with a few exceptions. Let $m$ be the total number of chains, $\tau$ the number of iterations per chain (where $\tau\geq2$), and $\theta$ the scalar summary of interest. For each chain ($\ell = 1, 2, \dots, m$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

<!--  [TODO: check this! isn't it because of under-dispersion and reliance on observed data?] It is unlikely that iterative algorithms such as MICE will achieve the ideal values of $\widehat{R}=1$ and $AC=0$. Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim, therefore, is to reach approximate convergence [@gelm13]. -->

Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} \gtrapprox 1$), and there is little dependency between subsequent iterations of imputation chains ($AC \gtrapprox 0$). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when $\widehat{R}$ and $AC$ exceed a certain threshold. The conventional thresholds to diagnose non-mixing are $\widehat{R} > 1.2$ [@gelm92] or $\widehat{R} > 1.1$ [@gelm13]. @vehtari2021 proposed a much more stringent threshold of $\widehat{R} > 1.01$. The magnitude of $AC$-values may be evaluated statistically, using a Wald test with $AC=0$ as null hypothesis [@box15]. $AC$-values that are significantly higher than zero indicate non-stationarity.
<!-- Threshold Rhat > 1.1 @gelm13, p. 288 -->

@@

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using the imputation package `{mice}` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{\mathrm{sim}} = 2000$). To evaluate convergence, we terminate the imputation algorithm after a varying number of iterations ($n_{\mathrm{it}} = 1, 2, ..., 100$, corresponding to $\tau$ in each individual imputation run). We differentiate between six different missingness scenarios that are defined in the data generating mechanism (see Table XYZ).

@@

<!-- [TODO: add guaranteed convergence because compatible imputation models, equivalent to JM and true Gibbs sampler, @vanbuuren2007] -->

We impute the missing data five times ($m = 5$) using Bayesian linear regression imputation with `mice` [@mice]. Bayesian linear regression is a compatible imputation model for the set of variables at hand. Therefore, the imputation algorithm is guaranteed to converge [@vanbuuren2007].

On each imputed dataset, we perform multiple linear regression as our analysis of scientific interest. Our estimands are the regression coefficients $\boldsymbol{\beta}$,
